---
title: "Lab 13 - Colonizing Mars"
author: "Barbara Mu"
date: "04/14/2026"
output: github_document
---

### Load packages and data

```{r load-packages, message=FALSE}
library(tidyverse) 

if (!require("ggplot2")) install.packages("ggplot2")
library(ggplot2)

if (!require("MASS")) install.packages("MASS")
library(MASS)

# Install and load the tidyverse package
if (!require("tidyverse")) install.packages("tidyverse")
library(tidyverse)
```

### Exercise 1

#### Exercise 1.1

```{r}
set.seed(123)
age <- rnorm(100, mean = 30, sd = 5)
```

```{r}
df_colonists <- data.frame(
  id = 1:100, # this is our id column
  age = age
) # add the age column here

head(df_colonists)
```

```{r}
df_colonists %>% 
  ggplot(aes(x = age)) + 
  geom_histogram(binwidth = 5) + 
  labs(x = "Age", 
       y = "Frequency")
```

The ages cluster around the population mean of 30, with most colonists falling between about 25 and 35 and a roughly symmetric, bell-shaped spread consistent with the `sd = 5` we specified. Because `rnorm()` is pseudo-random, the shape of any particular sample depends on the seed — `set.seed(123)` fixes the random stream so the histogram is reproducible. Changing the seed would shift which specific values appear (and tweak the skew of this small n = 100 sample), but the long-run distribution would still be approximately Normal(30, 5).

#### Exercise 1.2

```{r}
set.seed(1)
roleA <- rep(c("engineer", "scientist", "medic"),
  length.out = 100
)
# this works if we want to set a specific number of each role
roleB <- rep(c("engineer", "scientist", "medic"),
  each = 34,
  length.out = 100
)

roleC <- rep(c("engineer", "scientist", "medic"),
  times = c(33, 33, 33),
  length.out = 100
)

# if you want to use sampling weights
roleD <- sample(c("engineer", "scientist", "medic"),
  replace = TRUE,
  size = 100, prob = c(1, 1, 1)
)

# if you want to randomly shuffle

roleE <- sample(roleB, size = 100, replace = FALSE)
```


#### Exercise 1.3 

```{r}
df_colonists$role <- rep(c("engineer", "scientist", "medic"), length.out = 100)

table(df_colonists$role)
```

I went with `rep()` using `length.out = 100` to get a near-even split (34 engineers, 33 scientists, 33 medics). A young Mars colony needs all three roles simultaneously — engineers to keep habitats and life support running, scientists to do the exploratory/research work that justifies the mission, and medics to keep the crew alive. Balancing the roles roughly equally reflects the assumption that no single function can be deprioritized in the early phase; if the mission shifted (e.g., heavy construction push), I would reweight the sample using `prob =` in `sample()` instead.

#### Exercise 1.4

```{r}
set.seed(123)

df_colonists$marsgar <- runif(100, min = 0, max = 100)

summary(df_colonists$marsgar)
```

```{r}
df_colonists %>% 
  ggplot(aes(x = age, y = marsgar)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE)
```

### Exercise 2

#### Exercise 2.1

```{r}
# Simulate technical skills based on age

# Set the seed for reproducibility
set.seed(1235)

# Create a variable for technical skills
df_colonists$technical_skills <- 2 * df_colonists$age + rnorm(100, mean = 0, sd = 1)

summary(df_colonists$technical_skills)
```

```{r}
df_colonists %>% 
  ggplot(aes(x = age, y = technical_skills)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "Age", 
       y = "Technical Skills")
```

#### Exercise 2.2 

```{r}
# Simulate problem-solving abilities based on their assigned role

df_colonists$problem_solving[df_colonists$role == "engineer"] <- rnorm(sum(df_colonists$role == "engineer"), mean = 100, sd = 10)

df_colonists$problem_solving[df_colonists$role == "scientist"] <- rnorm(sum(df_colonists$role == "scientist"), mean = 80, sd = 10)

df_colonists$problem_solving[df_colonists$role == "medic"] <- rnorm(sum(df_colonists$role == "medic"), mean = 60, sd = 10)
```

```{r}
df_colonists %>%
  ggplot(aes(x = problem_solving, fill = role)) +
  geom_density(alpha = .5) + 
  labs(x = 'Problem-Solving Skills',
       y = 'Count',
       fill = 'Role') +
  scale_fill_viridis_d() +
  theme_minimal()
```


### Exercise 3

#### Exercise 3.1

```{r}
# Simulate three correlated traits: resilience, teamwork, adaptability
set.seed(7)

mean_traits <- c(50, 50, 50)

# Correlation matrix among the three traits
cor_traits <- matrix(c(
  1.00, 0.50, 0.40,
  0.50, 1.00, 0.35,
  0.40, 0.35, 1.00
), nrow = 3, byrow = TRUE)

# SDs of 10 on each trait → covariance matrix
sds <- c(10, 10, 10)
cov_traits <- diag(sds) %*% cor_traits %*% diag(sds)

traits_data <- mvrnorm(n = 100, mu = mean_traits, Sigma = cov_traits)

df_colonists$resilience <- traits_data[, 1]
df_colonists$teamwork <- traits_data[, 2]
df_colonists$adaptability <- traits_data[, 3]

# Check empirical correlations against the target matrix
round(cor(traits_data), 2)
```

#### Exercise 3.2

```{r}
# Pairwise scatter + correlations among the three traits
if (!require("GGally")) install.packages("GGally")
library(GGally)

df_colonists %>%
  dplyr::select(resilience, teamwork, adaptability) %>%
  ggpairs() +
  theme_minimal()
```

```{r}
# Correlation heatmap as an alternative view
trait_cor <- cor(df_colonists[, c("resilience", "teamwork", "adaptability")])

trait_cor_long <- as.data.frame(as.table(trait_cor))

ggplot(trait_cor_long, aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = round(Freq, 2)), color = "white") +
  scale_fill_viridis_c(limits = c(-1, 1), name = "r") +
  labs(title = "Correlations among simulated traits",
       x = NULL, y = NULL) +
  theme_minimal()
```

The scatterplot matrix and heatmap show positive correlations across all three traits, with resilience–teamwork (~0.50) being the strongest, matching the population values we fed into `mvrnorm()`.

#### Exercise 3.3

```{r}
cor_matrix_bigfive <- matrix(
  c(
1.0000, 0.2599, 0.1972, 0.1860, 0.2949,
0.2599, 1.0000, 0.1576, 0.2306, 0.0720,
0.1972, 0.1576, 1.0000, 0.2866, 0.1951,
0.1860, 0.2306, 0.2866, 1.0000, 0.1574,
0.2949, 0.0720, 0.1951, 0.1574, 1.0000
  ),
  nrow = 5, ncol = 5, byrow = TRUE,
  dimnames = list(
c("EX", "ES", "AG", "CO", "OP"),
c("EX", "ES", "AG", "CO", "OP")
  )
)
```

```{r}
big5_cor <- matrix(c(
#    O      C      E      A      N
  1.00,  0.27,  0.45,  0.28, -0.15,  # O
  0.27,  1.00,  0.26,  0.37, -0.40,  # C
  0.45,  0.26,  1.00,  0.25, -0.24,  # E
  0.28,  0.37,  0.25,  1.00, -0.29,  # A
 -0.15, -0.40, -0.24, -0.29,  1.00   # N
), ncol = 5)

colnames(big5_cor) <- rownames(big5_cor) <- 
  c("openness", "conscientiousness", "extraversion", 
    "agreeableness", "neuroticism")
```

```{r}
# Standard deviations for each trait
sds <- c(10, 10, 10, 10, 10)

# Convert correlation to covariance matrix
cov_big5 <- diag(sds) %*% big5_cor %*% diag(sds)
```

```{r}
set.seed(42)

# Means for all five traits
mean_big5 <- c(50, 50, 50, 50, 50)

# Generate correlated Big Five data
big5_data <- mvrnorm(n = 100,
                     mu = mean_big5,
                     Sigma = cov_big5)

# Add to dataframe
df_colonists$openness <- big5_data[, 1]
df_colonists$conscientiousness <- big5_data[, 2]
df_colonists$extraversion <- big5_data[, 3]
df_colonists$agreeableness2 <- big5_data[, 4]
df_colonists$neuroticism <- big5_data[, 5]

# Check how closely simulated correlations match population parameters
round(cor(big5_data), 2)
```

### Exercise 4

#### Exercise 4.1

```{r}
set.seed(42)

# Define a function for one colony simulation
simulate_colony <- function() {
  colony_data <- mvrnorm(n = 100,
                         mu = mean_big5,
                         Sigma = cov_big5)

  extrav <- colony_data[, 3]
  open <- colony_data[, 1]

  c(mean_extrav = mean(extrav),
    sd_extrav = sd(extrav),
    cor_extrav_open = cor(extrav, open))
}

# Run 100 times with replicate()
results <- replicate(100, simulate_colony())
results_df <- as.data.frame(t(results))

head(results_df)
```

```{r}
# Equivalent with a for-loop, for comparison
set.seed(42)

results_loop <- data.frame(
  mean_extrav = numeric(100),
  sd_extrav = numeric(100),
  cor_extrav_open = numeric(100)
)

for (i in 1:100) {
  colony_data <- mvrnorm(n = 100,
                         mu = mean_big5,
                         Sigma = cov_big5)

  results_loop$mean_extrav[i] <- mean(colony_data[, 3])
  results_loop$sd_extrav[i] <- sd(colony_data[, 3])
  results_loop$cor_extrav_open[i] <- cor(colony_data[, 3], colony_data[, 1])
}
```

#### Exercise 4.2

```{r}
pop_cor <- 0.45

ggplot(results_df, aes(x = cor_extrav_open)) +
  geom_histogram(binwidth = 0.05,
                 fill = "steelblue",
                 color = "white",
                 alpha = 0.8) +
  geom_vline(xintercept = pop_cor,
             color = "red",
             linetype = "dashed",
             linewidth = 1) +
  annotate("text",
           x = pop_cor + 0.02,
           y = 21,
           label = paste("Population r =", pop_cor),
           color = "red",
           hjust = 0) +
  labs(title = "Distribution of Extraversion–Openness Correlations Across 100 Colonies",
       x = "Correlation (r)",
       y = "Count") +
  theme_minimal()
```

```{r}
ggplot(results_df, aes(x = mean_extrav)) +
  geom_histogram(binwidth = 0.5,
                 fill = "purple",
                 color = "white",
                 alpha = 0.8) +
  geom_vline(xintercept = 50,
             color = "red",
             linetype = "dashed",
             linewidth = 1) +
  annotate("text",
           x = 51,
           y = 21,
           label = "Population mean = 50",
           color = "red",
           hjust = 0) +
  labs(title = "Distribution of Mean Extraversion Across 100 Colonies",
       x = "Mean Extraversion",
       y = "Count") +
  theme_minimal()
```

```{r}
ggplot(results_df, aes(x = sd_extrav)) +
  geom_histogram(binwidth = 0.25,
                 fill = "darkgreen",
                 color = "white",
                 alpha = 0.8) +
  geom_vline(xintercept = 10,
             color = "red",
             linetype = "dashed",
             linewidth = 1) +
  annotate("text",
           x = 10.2,
           y = 20,
           label = "Population SD = 10",
           color = "red",
           hjust = 0) +
  labs(title = "Distribution of Extraversion SD Across 100 Colonies",
       x = "SD of Extraversion",
       y = "Count") +
  theme_minimal()
```

Across 100 simulated colonies, the sample means for extraversion cluster tightly around the population mean of 50 and the sample SDs hover near the population SD of 10 — as expected from the central limit theorem with n = 100 per colony. The extraversion–openness correlations also center near the population value of 0.45 but show noticeably more spread, illustrating that correlation estimates are more variable than means at this sample size.

